---
title: "Course Project 1"
author: "Carlos Rubio"
date: "August 16, 2021"
output: html_document
keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data

Read and show summary of the data:

```{r}
data <- read.csv(file = 'activity.csv')
summary(data)
```

Histogram of number of steps per day

```{r}
groupedPerDay <- aggregate(data$steps, by=list(date=data$date), FUN=sum)
hist(groupedPerDay$x, main="Histogram of steps per day", xlab="Steps")
```

## What is mean total number of steps taken per day?

Mean and median of the total number of steps per day

```{r}
meanPerDay <- mean(groupedPerDay$x, na.rm=TRUE)
medianPerDay <- median(groupedPerDay$x, na.rm=TRUE)
cat("Mean steps per day: ", meanPerDay," steps\n")
cat("Median steps per day: ", medianPerDay," steps\n")
```

## What is the average daily activity pattern?

Group steps by interval and plot time series

```{r}
groupedByInterval <- aggregate(data$steps, by=list(interval=data$interval), FUN=mean, na.rm=TRUE)
plot(groupedByInterval$x, xlab="Interval index", ylab="Mean steps", main="Mean steps by 5-minute interval")
```

The 5-minute interval with maximum number of steps on average is:

```{r}
intervalMaxAverageSteps = groupedByInterval[which.max(groupedByInterval$x),]$interval
cat("Interval with max average steps: ", intervalMaxAverageSteps,"\n")
```

## Imputing missing values

Fill NA values with the mean for that 5-minute interval

```{r}
dataComplete <- data
incompleteIndexes = is.na(data$steps)
dataComplete$steps[incompleteIndexes] = groupedByInterval[groupedByInterval$interval == dataComplete$interval[incompleteIndexes],]$x
summary(dataComplete)
```

Histogram of number of steps per day in the complete dataset

```{r}
groupedPerDayComplete <- aggregate(dataComplete$steps, by=list(date=data$date), FUN=sum)
hist(groupedPerDayComplete$x, main="Histogram of steps per day (complete dataset)", xlab="Steps")
```

Mean and median of the total number of steps per day (complete dataset)

```{r}
meanPerDayComplete <- mean(groupedPerDayComplete$x, na.rm=TRUE)
medianPerDayComplete <- median(groupedPerDayComplete$x, na.rm=TRUE)
cat("Mean steps per day: ", meanPerDayComplete," steps\n")
cat("Median steps per day: ", medianPerDayComplete," steps\n")
```

## Are there differences in activity patterns between weekdays and weekends?

Encode week and week-end day in a factor variable

```{r}
dataComplete$day = factor(as.POSIXlt(dataComplete$date)$wday <= 5, c(TRUE, FALSE),c("weekday","weekend"))
```

Plot 5-minute interval time series

```{r fig.width = 12}
par(mfrow=c(1,2))
dataWeekday = dataComplete[dataComplete$day == "weekday",]
groupedByIntervalWeekday <- aggregate(dataWeekday$steps, by=list(interval=dataWeekday$interval), FUN=mean, na.rm=TRUE)
plot(groupedByIntervalWeekday$x, xlab="Interval index", ylab="Mean steps", main="Mean steps by 5-minute interval, weekdays")
dataWeekend = dataComplete[dataComplete$day == "weekend",]
groupedByIntervalWeekend <- aggregate(dataWeekend$steps, by=list(interval=dataWeekend$interval), FUN=mean, na.rm=TRUE)
plot(groupedByIntervalWeekend$x, xlab="Interval index", ylab="Mean steps", main="Mean steps by 5-minute interval, weekend")

```

